---
title: "Working with Spatial Data in R"
format: html
author: Huong Trinh
date: today
toc: true
toc-depth: 2
number-sections: false
theme: cosmo
---

## Set up

```{r}
# Load libraries
library(readr)
library(here)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
library(dplyr)
```

## Exploring the data using `plot()` and `st_crs()`

First, I read the Alaska region shapefile with `read_sf()`. I used `_sf` in the object name to remind myself this is spatial data.

```{r}
#| results: hide
#| fig-show: hide

# read the Alaska region shapefile
ak_rgns_sf <- read_sf(here("training_Htrinh", "data/ak_regions_simp.shp"))

# make a quick plot to see the data
plot(ak_rgns_sf)
```

I checked the object type and looked at the first rows. This helps me see what kind of data I have before doing more steps.

```{r}
#| results: hide

class(ak_rgns_sf)

head(ak_rgns_sf)

glimpse(ak_rgns_sf)
```

## Coordinate reference system

Spatial data need a coordinate reference system, also called CRS. The CRS tells R how the locations should be placed on the map.

```{r}
#| results: hide

st_crs(ak_rgns_sf)
```

The first map looked a little strange because the data were in longitude and latitude. I changed the Alaska region data to EPSG 3338 because this CRS is better for mapping Alaska.

```{r}
#| results: hide
#| fig-show: hide

ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

st_crs(ak_rgns_3338_sf)

plot(ak_rgns_3338_sf)
```

## `sf` and the tidyverse

An `sf` object is spatial data, but it can also work like a data frame. Because of this, I can use tidyverse functions like `select()` and `filter()`.

```{r}
#| results: hide

# check the column names in the data
colnames(ak_rgns_3338_sf)

ak_rgns_3338_sf %>%
    select(region)
```

```{r}
#| results: hide

unique(ak_rgns_3338_sf$region)

ak_rgns_3338_sf %>%
    filter(region == "Southeast")
```

## Spatial joins

The population data are in a CSV file. It has latitude and longitude columns, but R does not know it is spatial data yet.

```{r}
#| results: hide

# read the population CSV file
pop_df <- read_csv(here("training_Htrinh", "data/alaska_population.csv"))
```

I changed the population table into an `sf` object. I used longitude and latitude as the coordinates. I kept the original latitude and longitude columns because I need them later for the leaflet map.

```{r}
#| results: hide

pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),
                        crs = 4326,
                        remove = F)

head(pop_4326_sf)
```

The code below shows why CRS is important. This code is not run because the two spatial objects have different CRS, so it can make an error.

```{r}
#| eval: false

pop_joined_sf <- st_join(pop_4326_sf,
                         ak_rgns_3338_sf,
                         join = st_within)
```

To fix this problem, I changed the population points to the same CRS as the Alaska region polygons before joining them.

```{r}
#| results: hide

pop_3338_sf <- st_transform(pop_4326_sf,
                            crs = 3338)

pop_joined_sf <- st_join(pop_3338_sf,
                         ak_rgns_3338_sf,
                         join = st_within)

head(pop_joined_sf)
```

Now the joined data has the `region` column. This means each city point is matched with an Alaska region.

```{r}
#| results: hide

pop_joined_sf %>%
    st_drop_geometry() %>%
    select(city, population, region, mgmt_area) %>%
    head()
```

## Calculate total population by region

Next, I calculated the total population for each region. I dropped the geometry first because this part is only a table summary.

```{r}
#| results: hide

pop_rgn_df <- pop_joined_sf %>%
    st_drop_geometry() %>%
    group_by(region) %>%
    summarize(total_pop = sum(population))

head(pop_rgn_df)
```

Then I joined the population summary back to the Alaska region polygons. This gives me spatial polygons with population values, so I can make a map.

```{r}
#| results: hide
#| fig-show: hide

pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf,
                             pop_rgn_df,
                             by = "region")

# make a quick plot to check the population data
plot(pop_rgn_3338_sf["total_pop"])
```

## Summarize by management area

This code summarizes the population by management area. In this first version, the regions are combined by management area.

```{r}
#| results: hide
#| fig-show: hide

pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop))

plot(pop_mgmt_3338_sf["total_pop"])
```

In this second version, I used `do_union = F` so the shapes are not combined into one big shape.

```{r}
#| results: hide
#| fig-show: hide

pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop), do_union = F)

plot(pop_mgmt_3338_sf["total_pop"])
```

## Save the spatial object

This code can save the new spatial object as a shapefile. I kept it commented out because I do not need to create a new file every time I render the document.

```{r}
#| results: hide

write_sf(pop_rgn_3338_sf, here("training_Htrinh", "data", "ak_regions_population.shp"))
```

## Visualize with `ggplot`

This code makes a simple population map. I hid this map in the HTML because I only want to show the final map later.

```{r}
#| fig-show: hide

ggplot(pop_rgn_3338_sf) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high = "firebrick",
                          labels = comma) +
    theme_bw()
```

For the final static map, I added the Alaska regions, city points, and rivers together.

```{r}
rivers_3338_sf <- read_sf(here("training_Htrinh", "data/ak_rivers_simp.shp"))

st_crs(rivers_3338_sf)
```

```{r}
ggplot() +
    geom_sf(data = pop_rgn_3338_sf,
            aes(fill = total_pop)) +
    geom_sf(data = pop_3338_sf,
            size = 0.5) +
    geom_sf(data = rivers_3338_sf,
            aes(linewidth = StrOrder)) +
    scale_linewidth(range = c(0.05, 0.5),
                    guide = "none") +
    labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high = "firebrick",
                          labels = comma) +
    theme_bw()
```

## Optional basemap with `ggspatial`

This part is optional. I kept the code here, but I did not run it because online map tiles can take more time and this part is not required.

```{r}
#| eval: false

pop_3857_sf <- st_transform(pop_3338_sf,
                            crs = 3857)

ggplot(data = pop_3857_sf) +
    ggspatial::annotation_map_tile(type = "osm", zoom = 4, progress = 'none') +
    # a bigger zoom number gives more map detail
    geom_sf(aes(color = population),
            fill = NA) +
    scale_color_continuous(low = "darkkhaki",
                           high = "firebrick",
                           labels = comma)
```

## Visualize `sf` objects with `leaflet`

For leaflet, I changed the region population data back to WGS84. Leaflet works better with longitude and latitude coordinates.

```{r}
#| results: hide

epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def = "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0
        +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)

st_crs(pop_rgn_3338_sf)

pop_rgn_4326_sf <- pop_rgn_3338_sf %>%
    st_transform(crs = 4326)
```

This first leaflet map is a simple version. I kept it as code only so the HTML does not show repeated maps.

```{r}
#| eval: false

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(data = pop_rgn_4326_sf,
                fillColor = "gray",
                weight = 1)

m
```

This second leaflet map adds color, labels, and a legend. I also kept it as code only because I only want the final interactive map to show.

```{r}
#| eval: false

pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~ region
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

m
```

The final interactive map shows the Alaska regions and the city points. The city labels show the city name and population.

```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1
    ) %>%
    addCircleMarkers(
        data = pop_4326_sf,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500),
        # this makes bigger populations show as bigger circles
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326_sf$city, ", population ",
                         comma(pop_4326_sf$population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

m
```

## Reflection

This assignment helped me understand why CRS is very important in spatial data. The spatial join did not work until the population points and Alaska region polygons used the same CRS. After I joined the points to the regions, I could summarize the population by region and make both a static map and an interactive map.
